thesis2024

### 2025-05-06 06-discussion.qmd
Project organised, all articles and chapters compilled and revised

In [@sec-chapter1], we outlined the research questions essential to achieving the overall goals of this doctoral dissertation and project. The answers to these questions collectively provide a detailed picture of how the Upper Paleolithic (UP) hunter-gatherers of southwestern Iberia obtained and managed their resources, moved across the landscape, and connected with other regions of Iberia, despite their peripheral location.

In this chapter, we address each of those research questions, through the integration and synthesis of all the interpretations and conclusions presented in the previous chapters.

What types of cherts were exploited and how were they introduced into the site?

Throughout the UP sequence at Vale Boi, hunter-gatherers exploited a wide variety of cherts. Although this research primarily focuses on chert, it is important to note that chert constituted only a portion of the lithic raw materials used at the site. Other materials—such as quartz and greywacke—were also intensively exploited, often for specific activities (Cascalheira et al., 2017; Pereira et al., 2016; Horta et al., 2019).

@@

No chert from more distant regional sources (e.g., \>80 km to the east) was identified in the UP sequence. This suggests that, despite the broader availability of chert throughout the Algarve region, the hunter-gatherer groups at Vale Boi exploited only a limited portion of the region’s lithic resources. Based on current geological and archaeological evidence—as well as indications of coastal resource foraging throughout the UP (Manne & Bicho, 2011)—it appears that the foraging radius of these groups during occupations at Vale Boi was approximately 20 km (sensu Roebroeks et al., 1988, cited in Brantingham, 2003). Chert procurement was likely embedded within broader foraging strategies (Binford, 1980).

A portion of the cherts used at Vale Boi were non-local. In most cases, these non-local chert types are consistently present across all occupation levels and areas—excluding T11, which appears only in the Solutrean occupations, and the TL cherts, which are sporadic and represented by small groups or individual artifacts of distinct macroscopic lithotypes.

At least two different transportation strategies were identified throughout the UP sequence at Vale Boi. The majority of non-local cherts appear to have been transported to the site as nodules, partially decortified cores, or larger stone volumes, and subsequently knapped on-site. This is evidenced by the consistent presence of cortex across all non-local chert types—though notably lacking the rounded, pebble-like cortex typical of beach or river sources—as well as by the presence of cores and associated knapping by-products. These findings suggest a transportation strategy focused on larger volumes of raw material, as proposed in previous studies (Clark 1987; Gould et al. 1971; Shott 1986; Kelly 1988; Perlès 1992; Kuhn 1994; Morrow 1996), and are indicative of a site provisioning approach (Kuhn, 2004).

A similar model of long-distance raw material transportation is documented in France, where in the Late Gravettian hunter-gatherers transported complete or minimally pre-formed blocks over distances exceeding 200 km, in preparation for short-term hunting occupations (Delvigne et al., 2021).

By comparison, the Gravettian occupations of the Côa Valley exhibit intermediate patterns of long-distance raw material use. While some portion of the non-local resources arrived as nodules—and are found in the archaeological record as reduced cores, core preparation, and maintenance products indicating on-site knapping—a significant portion of the imported raw materials shows reduction sequences corresponding to the final phases of bladelet production and the reuse of blanks (Aubry et al., 2012).

This contrasts with patterns observed at other UP sites across Iberia in relation to long-distance chert procurement. For example, at the Ametzagaina site in the western Pyrenees (Basque Country), non-local raw materials from sources located approximately 50–100 km away are present in Gravettian occupations. These materials are represented mainly by retouched tools and burin spalls, with no associated cores, although core preparation and maintenance tools are present (Arrizabalaga et al., 2014).

More similar to these finished-product transports, a small portion of non-local cherts at Vale Boi—possibly transported as blanks or retouched tools—was also identified throughout the UP sequence. These are represented in the TL cherts and may reflect provisioning strategies aligned with individual provisioning, as proposed by Kuhn (2004), involving the transport of blanks and tools as part of an individual's portable toolkit.

@@

-   Cherts from the Cretaceous formations of central Portugal, found near the Lisbon and Rio Maior region (Matias 2016; Matias 2012; Jordão and Pimentel 2022), located approximately \~215-250 km north of Vale Boi (as the crow flies). The attribution is currently based on macroscopic, petrographic and chemical similarities between archaeological and geological samples.

-   Cherts from the Upper Jurassic formations of the Betic Systems in southern Spain (Rodríguez et al. 2011), located over \~250 km east of Vale Boi (as the crow flies). The identification is based primarily on macroscopic traits and petrographic components not found in local or regional cherts.

-   Although much less represented, the presence of jasper in the TL chert group and the distinct geochemical signature of T11, as identified through pXRF analysis, suggest these artefacts may originate from the Iberian Pyritic Belt (Araújo et al., 2013; Oliveira et al., 1991), located north of the Algarve region. While these resources are not available in primary deposits in the Algarve, they are known to occur in secondary deposits, such as beaches or riverbeds, which could have brought them closer to the site.

@@

How frequently did these groups move across the landscape, and can long-distance movements or social contacts be identified through the presence of non-local cherts?

@@

While the UP occupations at Vale Boi have been interpreted as the result of seasonal mobility, possibly linked to deer hunting activities in spring (Manne et al., 2012), we do not argue that the presence of cherts from over 200 km away indicates a single cultural group occupying the entire southern half of Iberia. Nor does the evidence support the idea that these materials arrived at the site solely through direct or embedded lithic procurement across such vast distances.

@@

Individually, these residential moves are feasible. Several ethnographic examples document total annual residential movements between 400 and 500 km, such as those of the Mistassini Cree (Rogers, 1963), Tasmanians (Jones, 1974), and Nukak (Politis, 2007). However, such mobility often involves the use of aids like boats or sleds. In a broader synthesis of hunter-gatherer annual mobility, Kelly (1995) shows that only under extreme environmental conditions—such as those faced by the Baffinland Inuit (Hantzsch, 1977) or the Nunamiut (Binford, 1978)—do residential movements approach the combined distances inferred for Vale Boi when considering the northern and eastern trajectories together.

However, these distances are calculated as straight-line measurements from Vale Boi to the raw material sources, without accounting for topographic or environmental factors. Long-distance movements in those directions may have been shaped by terrain constraints, a factor incorporated into formal models of hunter-gatherer mobility (Brantingham, 2006).

When applied to the Iberian Peninsula, terrain friction mapping demonstrates that travel costs on foot vary significantly from direct distances measured on modern maps (Díaz del Río, 2020). In this context, distance is better understood in terms of travel time rather than absolute spatial measurements (Zilhão, 2021). Díaz del Río (2020) highlights the presence of low-cost corridors—particularly across plateaus and along the interconnected river valleys of central and southern Iberia, including the Guadalquivir, Guadiana, and Tagus. This suggests that the apparent long distances from Vale Boi to chert sources in central Portugal and southern Spain may have been less arduous and time-consuming than they seem on a flat map.

This may also be reflected in other long-distance movements identified in western Europe. In France, archaeological evidence from Gravettian seasonal hunting camps (e.g., Le Blot and Le Rond-de-Saint-Arçon) shows the exclusive use of non-local raw materials, indicating their integration into large mobility circuits within broader territorial ranges (Delvigne et al., 2021).

@@

However, several technological studies of UP lithic assemblages at Vale Boi have revealed key differences in knapping strategies, raw material use, and typology when compared to assemblages from central Portugal and southern Spain (Marreiros et al., 2015). Vale Boi has often been characterized as a site reflecting resilient behaviours and cultural continuity (Manne et al., 2012; Cascalheira et al., 2017), but also, particularly during the Gravettian, as exhibiting elements of cultural distinction from neighbouring groups (Marreiros et al., 2015).

While some of these differences may reflect local environmental adaptations or the specific functional characteristics of the site, the distinct typological choices—often linked to style and cultural identity—strongly suggest that the Gravettian hunter-gatherers at Vale Boi constituted a separate cultural group from those in central Portugal and southern Spain. Nevertheless, this cultural distinction does not preclude the possibility of interaction. This scenario aligns with Steward’s (1969) concept of broader social networks, or the "maximum band," defining them as a group with which its members (the smaller bands of groups) vaguely identify.

Regarding raw material procurement, while it is not unheard of in ethnographic examples for one group to access the territory of another—often with permission and under established social agreements (Jones, 1974; Kelly, 1995)—distances greater than 200 km for resource acquisition are rarely documented. Ethnographic data generally show that residential mobility seldom exceeds 300 km annually (Kelly, 1995), and that raw materials procured during an occupation are typically sourced within a 20 km radius (Binford, 2001). Materials acquired from beyond this range are more often the result of transport from previous locations, rather than direct procurement during a single occupation (Surovell, 2009).

@@

Archaeological examples in western Europe document levels of mobility comparable to those observed at Vale Boi, with many interpretations emphasizing the role of social networks connecting hunter-gatherer groups.

For instance, UP sites in the Côa Valley (inland northern Portugal) exhibit patterns of long-distance—or "exotic"—chert acquisition strikingly similar to those at Vale Boi. Spanning the Gravettian to the Magdalenian (though with poorly preserved Solutrean occupations), the Côa Valley sites reveal non-local cherts sourced from the Rio Maior region (southwest; 130–225 km) and the Spanish Northern Meseta (east; 110–205 km). These materials have been interpreted as evidence of inter-group exchanges and the existence of social networks, with contact likely occurring along territorial boundaries during periods of aggregation—a pattern well-documented among hunter-gatherer societies (Aubry et al., 2012; 2015; 2016; 2022).

This interpretation is further supported by other studies linking sites in southern France (e.g., the Paris Basin) and central Portugal. During the Solutrean and Last Glacial Maximum, these sites show similar raw material procurement and technological patterns over local and regional distances, suggesting a broad interaction network spanning southwestern Iberia and western France—even during periods like the Solutrean, which are poorly represented in the Côa Valley (Aubry et al., 2003).

Additional evidence comes from northern Iberia. At the Valverde site in Cantabria, non-local cherts sourced from over 190 km away were identified in Solutrean levels. Although rare, their presence supports the idea that UP groups in Galicia participated in an extensive network of social contacts stretching across northern Iberia and into France (Hermida et al., 2016).

Similarly, the Peña Capón site in the Spanish Southern Meseta shows varying percentages of non-local cherts throughout the UP sequence, with the highest quantities during the Solutrean and decreasing presence in earlier periods like the Proto-Solutrean and Gravettian. Macroscopic analysis suggests sources located over 150 km away, interpreted as evidence for contact or mobility between the Tagus and Ebro valleys (Sánchez de la Torre et al., 2023).

Given these parallels, the presence of long-distance raw materials at Vale Boi can plausibly be interpreted as evidence of social interaction—possibly facilitated during seasonal aggregations (Lee, 1972), at the margins of distinct territories (Brantingham, 2006), or within overlapping mating networks (Wobst, 1976), all contributing to the reinforcement of social bonds.

Ethnographic studies further underscore the importance of social networks in explaining long-distance mobility, even when residential mobility remains limited. Long-distance movement often served key adaptive purposes, enabling the formation and maintenance of broader social ties (Gould & Saggers, 1985). While a minimum band could subsist in relative isolation, multiple bands commonly participated in a larger social network to improve survival prospects and preserve cultural continuity (Wobst, 1974). This pattern may have been especially critical in marginal regions, where social exchange—including mating partnerships, rituals, symbolic systems, and resource trade—was essential for sustaining viable communities (Wobst, 1976).

Ethnographic studies demonstrate that expansive social networks are often developed by hunter-gatherers as adaptive responses to risk, especially in environments characterized by irregular and unpredictable resource availability (Gould, 1978). In such contexts, social relationships serve as mechanisms to mitigate scarcity through information exchange, mutual support, and the circulation of resources across vast areas.

For example, ethnographic accounts of !Kung hunter-gatherers show that groups or individuals travelled long distances to gather for winter-gathering, or even for long-distance exchange networks. In their exchanges, goods travelled across hundreds of miles from individual to individual (Lee, 1972). These exchanges were not limited to utilitarian needs but played a vital role in maintaining inter-group ties. Within this framework, the acquisition of exotic goods, including lithic raw materials, can be understood as a byproduct of social cohesion rather than procurement (Wobst, 1974; Gould & Saggers, 1985).

By interpreting the presence of non-local resources at Vale Boi as the result of exchange and social networks, then we assume the existence of at least two maintained social networks, previously suggested by technological studies (focused mainly in the Proto-Solutrean and Solutrean technocomplexes; Belmiro et al., 2021; Cascalheira, 2013).

However, the existence of long-distance mobility in hunter-gatherer groups during the UP is, in most cases, expected. Except for situations where specific technological choices may be interpreted as ethnographic and cultural boundaries between different groups, possibly indicating small territories, the very basis of a hunter-gatherer lifeway is marked by mobility and adaptation to the shifts in natural resources and the necessity to respect their renewal times (Kelly, 1995).

@@

Evidence from the site suggests that residential mobility patterns evolved throughout the UP. During the Gravettian, Vale Boi presents archaeological signals of short-term occupations, indicating higher mobility. This correlates with greater frequencies of non-local cherts, particularly those in the TL group, which may reflect either broader ranging mobility or individual provisioning strategies. In contrast, from the Proto-Solutrean onward, Vale Boi transitions into a longer-term residential settlement. Correspondingly, the frequency of non-local lithics declines, with local cherts becoming dominant during the Solutrean, consistent with reduced residential mobility (Surovell, 2009; Grove, 2023).

How were different chert types managed by hunter-gatherer groups, and what impact did mobility have on their technological organisation?

These shifts in mobility frequency during the UP not only influenced the quantity of non-local raw materials found at Vale Boi but also shaped the technological and typological characteristics of the lithic assemblages. When focusing specifically on the relationship between occupation duration and technological behavior, short-term occupations—associated with higher mobility—tend to correspond with increased frequencies of retouched tools, a pattern well-documented in ethnographic and archaeological literature (Torrence 1983; Surovell 2009; Grove et al. 2023). The more frequent occurrence of burins, for instance, may reflect an adaptive strategy to mobility, as these tools could serve as cores for the production of bladelets (Barton et al. 1996).

Across all UP phases at Vale Boi, regardless of occupation length, non-local cherts—costlier in terms of acquisition time and transport effort, even when acquired through exchange—consistently show higher proportions of retouched tools and distinct tool types. This pattern supports interpretations that non-local materials were not only more carefully curated but also reflect a higher investment in production and maintenance, aligning with theoretical expectations surrounding raw material economy (Kelly 1988; Perlès 1992; Andrefsky 1994; Kuhn 2004).

The types of produced tool show diverse set of tool types, although less specialized tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages. This suggests that these materials were not only transported as unspecialised tools and easily reconfigurable tools (e.g., retouched blanks) as suggested in previous studies (Odell 1981; Siegel 1984; Shott 1986), but as volumes that supported tool specialisation and diversification.

In addition, non-local cherts consistently exhibit higher knapping intensity, indicating an increased effort to maximize their utility once transported to the site. This behavior conforms to both theoretical models and archaeological expectations regarding the conservation and extended use of high-cost raw materials (Roth and Dibble 1998; Kuhn 2004; Surovell 2009).

This clearly indicates that despite variations in residential mobility throughout the UP, the lithic record at Vale Boi reveals differentiated strategies of raw material management, where distance from source emerges as a key determinant in shaping technological investment and tool use.

How did technological and social organisation evolve over approximately 10,000 years of the UP in southwestern Portugal, and how might climatic changes have influenced these adaptations?

@@

After the Heinrich stadial, which once again created mild, mediterranean conditions all throughout Iberia, the onset of the Proto-Solutrean and, subsequently, the Solutrean, coincided with Heinrich Event 2 and the Last Glacial Maximum (LGM) (Belmiro et al., 2021; Cascalheira and Bicho, 2013; González-Sampériz et al., 2010; Schmidt et al., 2012). It is also during this time frame (c. 26 and 20 ka cal BP; Belmiro et al. 2021; Cascalheira et al. 2017; Cascalheira and Bicho 2013) that the settlement patterns and raw material proportions begin to show alterations, after what may have been a small occupational hiatus before the Proto-Solutrean represented by lower frequencies of archaeological materials.

As such, we suggest that as climatic conditions worsened, transforming the southern Iberian landscape from Mediterranean forests to desert-like steppes (Gómez and Lunt, 2006; Gonzalez-Sampériz et al., 2010; Hewitt, 2000), hunter-gatherer groups continued seasonal mobility, but at a reduced frequency. According to Surovell’s (2009) “Mean per Capita Occupation Span” model, this reduction likely stemmed from resource scarcity and the need to avoid overexploitation of ecological niches.

A similar pattern is observed in the UP occupation of the Côa Valley, where from the Proto-Solutrean onwards, the proportions of non-local raw materials decreased with an increased reliance on regional resources in comparison to the Gravettian occupations (Aubry et al., 2012). This may indicate a similar behaviour across Iberia motivated by the occupation of ecological niches.

While it would then be expected that social networks would see themselves expanded during the LGM given the scarcity of resources, and thus a higher presence of long-distance raw materials on Proto-Solutrean and Solutrean occupations, the existence of these ecological niches instead possibly fostered an environment of relative stability.

>>>>

In [@sec-chapter1], we outlined the research questions essential to achieving the overall goals of this doctoral dissertation and project. The answers to these questions collectively provide a detailed picture of how the Upper Palaeolithic (UP) hunter-gatherers of southwestern Iberia obtained and managed their resources, moved across the landscape, and connected with other regions of Iberia, despite their peripheral location.

In this chapter, we address those research questions by integrating and synthesising all the interpretations and conclusions presented in the previous chapters.

*What types of cherts were exploited, and how were they introduced into the site?*

Throughout the UP sequence at Vale Boi, hunter-gatherers exploited various cherts. Although this research primarily focuses on chert, it is important to note that chert constituted only a portion of the lithic raw materials used at the site. Other materials—such as quartz and greywacke—were also intensively exploited, often for specific activities [@cascalheira_2017; @pereira_territory_2016-1; @horta_role_2019].

@@

No chert from more distant regional sources (e.g., \>80 km to the east) was identified in the UP sequence. This suggests that, despite the broader availability of chert throughout the Algarve region, the hunter-gatherer groups at Vale Boi exploited only a limited portion of the region’s lithic resources. Based on current geological and archaeological evidence—as well as indications of coastal resource foraging throughout the UP [@mannePryingNewMeaning2011]—it appears that the foraging radius of these groups during occupations at Vale Boi was approximately 20 km [sensu Roebroeks et al., 1988, cited in @brantinghamNeutralModelStone2003]. Chert procurement was likely embedded within broader foraging strategies [@binfordWillowSmokeDogs1980].

A portion of the cherts used at Vale Boi were non-local. In most cases, these non-local chert types are consistently present across all occupation levels and areas, excluding T11, which appears only in the Solutrean occupations, and the TL cherts, which are sporadic and represented by small groups or individual artefacts of distinct macroscopic lithotypes.

At least two different transportation strategies were identified throughout the UP sequence at Vale Boi. The majority of non-local cherts appear to have been transported to the site as nodules, partially decortified cores, or larger stone volumes, and subsequently knapped on-site. This is evidenced by the consistent presence of cortex across all non-local chert types—though notably lacking the rounded, pebble-like cortex typical of beach or river sources—as well as by the presence of cores and associated knapping by-products. These findings suggest a transportation strategy focused on larger volumes of raw material, as proposed in previous studies [@clarkPoliticsPrismaticBlades1987; @gouldLithicAssemblageWestern1971; @shottTechnologicalOrganizationSettlement1986; @kellyThreeSidesBiface1988; @perlesSearchLithicStrategies1992; @kuhnFormalApproachDesign1994; @morrowBiggerBetterComments1996], and are indicative of a site provisioning approach [@kuhnUpperPaleolithicRaw2004].

A similar model of long-distance raw material transportation is documented in France, where in the Late Gravettian hunter-gatherers transported complete or minimally pre-formed blocks over distances exceeding 200 km, in preparation for short-term hunting occupations [@delvigneGeoresourcesTechnoculturalExpressions2019].

By comparison, the Gravettian occupations of the Côa Valley exhibit intermediate long-distance raw material use patterns. While some portion of the non-local resources arrived as nodules—and are found in the archaeological record as reduced cores, core preparation, and maintenance products indicating on-site knapping—a significant portion of the imported raw materials shows reduction sequences corresponding to the final phases of bladelet production and the reuse of blanks [@aubryWeWillBe2012].

This contrasts with patterns observed at other UP sites across Iberia in relation to long-distance chert procurement. For example, at the Ametzagaina site in the western Pyrenees (Basque Country), non-local raw materials from sources located approximately 50–100 km away are present in Gravettian occupations. These materials are represented mainly by retouched tools and burin spalls, with no associated cores, although core preparation and maintenance tools are present [@arrizabalagaWhereWhat2014].

More similar to these finished-product transports, a small portion of non-local cherts at Vale Boi—possibly transported as blanks or retouched tools—was also identified throughout the UP sequence. These are represented in the TL cherts and may reflect provisioning strategies aligned with individual provisioning, as proposed by Kuhn [-@kuhnUpperPaleolithicRaw2004], involving the transport of blanks and tools as part of an individual's portable toolkit.

@@

-   Cherts from the Cretaceous formations of central Portugal, found near the Lisbon and Rio Maior region [@matiasAprovisionamentoMateriasprimasLiticas2012; @matiasRawMaterialSourcing2016; @jordaoFlintSourcesMobility2022], located approximately \~215-250 km north of Vale Boi (as the crow flies). The attribution is currently based on macroscopic, petrographic and chemical similarities between archaeological and geological samples.

-   Cherts from the Upper Jurassic formations of the Betic Systems in southern Spain [@rodriguezPrehistoricFlintExploitations2011], located over \~250 km east of Vale Boi (as the crow flies). The identification is based primarily on macroscopic traits and petrographic components not found in local or regional cherts.

-   Although much less represented, the presence of jasper in the TL chert group and the distinct geochemical signature of T11, as identified through pXRF analysis, suggest these artefacts may originate from the Iberian Pyritic Belt [@araujoRegioesCentralSul2013; @oliveiraTracosGeraisEvolucao1991], located north of the Algarve region. While these resources are not available in primary deposits in the Algarve, they are known to occur in secondary deposits, such as beaches or riverbeds, which could have brought them closer to the site.

@@

*How frequently did these groups move across the landscape, and can long-distance movements or social contacts be identified through the presence of non-local cherts?*

@@

While the UP occupations at Vale Boi have been interpreted as the result of seasonal mobility, possibly linked to deer hunting activities in spring [@manneIntensiveSubsistencePractices2012], we do not argue that the presence of cherts from over 200 km away indicates a single cultural group occupying the entire southern half of Iberia. Nor does the evidence support the idea that these materials arrived at the site solely through direct or embedded lithic procurement across such vast distances.

@@

Individually, these residential moves are feasible. Several ethnographic examples document total annual residential movements between 400 and 500 km, such as those of the Mistassini Cree [@rogersChangingSettlementPatterns1963], Tasmanians [@jonesTasmanianTribes1974], and Nukak [@politisNUKAKEthnoarchaeologyAmazonian2007]. However, such mobility often involves using aids like boats or sledges. In a broader synthesis of hunter-gatherer annual mobility, Kelly [-@kellyForagingSpectrumDiversity1995] shows that only under extreme environmental conditions—such as those faced by the Baffinland Inuit [@hantzschMyLifeEskimos1977] or the Nunamiut [@binfordNunamiutEthnoarchaeology1978]—do residential movements approach the combined distances inferred for Vale Boi when considering the northern and eastern trajectories together.

However, these distances are calculated as straight-line measurements from Vale Boi to the raw material sources, without accounting for topographic or environmental factors. Long-distance movements in those directions may have been shaped by terrain constraints, a factor incorporated into formal models of hunter-gatherer mobility [@brantinghamMeasuringForagerMobility2006].

When applied to the Iberian Peninsula, terrain friction mapping demonstrates that travel costs on foot vary significantly from direct distances measured on modern maps [@diazdelrioWhatIberianCopper2020]. In this context, distance is better understood in terms of travel time rather than absolute spatial measurements [@zilhaoLatePersistenceMiddle2021]. Díaz del Río [-@diazdelrioWhatIberianCopper2020] highlights the presence of low-cost corridors—particularly across plateaus and along the interconnected river valleys of central and southern Iberia, including the Guadalquivir, Guadiana, and Tagus. This suggests that the apparent long distances from Vale Boi to chert sources in central Portugal and southern Spain may have been less arduous and time-consuming than they seem on a flat map.

This may also be reflected in other long-distance movements identified in western Europe. In France, archaeological evidence from Gravettian seasonal hunting camps (e.g., Le Blot and Le Rond-de-Saint-Arçon) shows the exclusive use of non-local raw materials, indicating their integration into large mobility circuits within broader territorial ranges [@delvigneGeoresourcesTechnoculturalExpressions2019].

@@

However, several technological studies of UP lithic assemblages at Vale Boi have revealed key differences in knapping strategies, raw material use, and typology when compared to assemblages from central Portugal and southern Spain [@marreirosLithicTechnologyGravettian2015]. Vale Boi has often been characterised as a site reflecting resilient behaviours and cultural continuity [@manneIntensiveSubsistencePractices2012; @cascalheira_2017], but also, particularly during the Gravettian, as exhibiting elements of cultural distinction from neighbouring groups [@marreirosLithicTechnologyGravettian2015].

While some of these differences may reflect local environmental adaptations or the specific functional characteristics of the site, the distinct typological choices—often linked to style and cultural identity—strongly suggest that the Gravettian hunter-gatherers at Vale Boi constituted a separate cultural group from those in central Portugal and southern Spain. Nevertheless, this cultural distinction does not preclude the possibility of interaction. This scenario aligns with Steward’s [-@stewardPostscriptBandsTaxonomy1969] concept of broader social networks, or the "maximum band," defining them as a group with which its members (the smaller bands of groups) vaguely identify.

Regarding raw material procurement, while it is not unheard of in ethnographic examples for one group to access the territory of another—often with permission and under established social agreements [@jonesTasmanianTribes1974; @kellyForagingSpectrumDiversity1995]—distances greater than 200 km for resource acquisition are rarely documented. Ethnographic data generally show that residential mobility seldom exceeds 300 km annually [@kellyForagingSpectrumDiversity1995], and that raw materials procured during an occupation are typically sourced within a 20 km radius [@binfordConstructingFramesReference2001]. Materials acquired from beyond this range are more often the result of transport from previous locations, rather than direct procurement during a single occupation [@surovellBehavioralEcologyLithic2009].

@@

Archaeological examples in western Europe document levels of mobility comparable to those observed at Vale Boi, with many interpretations emphasising the role of social networks connecting hunter-gatherer groups.

For instance, UP sites in the Côa Valley (inland northern Portugal) exhibit patterns of long-distance—or "exotic"—chert acquisition strikingly similar to those at Vale Boi. Spanning the Gravettian to the Magdalenian (though with poorly preserved Solutrean occupations), the Côa Valley sites reveal non-local cherts sourced from the Rio Maior region (southwest; 130–225 km) and the Spanish Northern Meseta (east; 110–205 km). These materials have been interpreted as evidence of inter-group exchanges and the existence of social networks, with contact likely occurring along territorial boundaries during periods of aggregation—a pattern well-documented among hunter-gatherer societies [@aubryFarFlintInferring2022; @aubryWeWillBe2012; @aubryUpperPalaeolithicLithic2016; @aubryAdaptationResourcesEnvironments2015].

This interpretation is further supported by other studies linking sites in southern France (e.g., the Paris Basin) and central Portugal. During the Solutrean and Last Glacial Maximum, these sites show similar raw material procurement and technological patterns over local and regional distances, suggesting a broad interaction network spanning southwestern Iberia and western France—even during periods like the Solutrean, which are poorly represented in the Côa Valley [@aubrySolutreanLaurelLeaf2003].

Additional evidence comes from northern Iberia. At the Valverde site in Cantabria, non-local cherts sourced from over 190 km away were identified in Solutrean levels. Although rare, their presence supports the idea that UP groups in Galicia participated in an extensive network of social contacts stretching across northern Iberia and into France [@hermidaSilexNWPeninsula2016].

Similarly, the Peña Capón site in the Spanish Southern Meseta shows varying percentages of non-local cherts throughout the UP sequence, with the highest quantities during the Solutrean and decreasing presence in earlier periods like the Proto-Solutrean and Gravettian. Macroscopic analysis suggests sources located over 150 km away, interpreted as evidence for contact or mobility between the Tagus and Ebro valleys [@sanchezdelatorreNewDataChert2023].

Given these parallels, the presence of long-distance raw materials at Vale Boi can plausibly be interpreted as evidence of social interaction—possibly facilitated during seasonal aggregations [@leeIntensificationSocialLife1972], at the margins of distinct territories [@brantinghamMeasuringForagerMobility2006], or within overlapping mating networks [@wobstLocationalRelationshipsPaleolithic1976], all contributing to the reinforcement of social bonds.

Ethnographic studies further underscore the importance of social networks in explaining long-distance mobility, even when residential mobility remains limited. Long-distance movement often served key adaptive purposes, enabling the formation and maintenance of broader social ties [@gouldEmpiricistStrikesBack1985]. While a minimum band could subsist in relative isolation, multiple bands commonly participated in a larger social network to improve survival prospects and preserve cultural continuity [@wobstBoundaryConditionsPaleolithic1974]. This pattern may have been especially critical in marginal regions, where social exchange—including mating partnerships, rituals, symbolic systems, and resource trade—was essential for sustaining viable communities [@wobstLocationalRelationshipsPaleolithic1976].

Ethnographic studies demonstrate that expansive social networks are often developed by hunter-gatherers as adaptive responses to risk, especially in environments characterised by irregular and unpredictable resource availability [@gouldAnthropologyHumanResidues1978]. In such contexts, social relationships serve as mechanisms to mitigate scarcity through information exchange, mutual support, and the circulation of resources across vast areas.

For example, ethnographic accounts of !Kung hunter-gatherers show that groups or individuals travelled long distances to gather for winter-gathering, or even for long-distance exchange networks. In their exchanges, goods travelled across hundreds of miles from individual to individual [@leeIntensificationSocialLife1972]. These exchanges were not limited to utilitarian needs but played a vital role in maintaining inter-group ties. Within this framework, the acquisition of exotic goods, including lithic raw materials, can be understood as a byproduct of social cohesion rather than procurement [@wobstBoundaryConditionsPaleolithic1974; @gouldEmpiricistStrikesBack1985].

By interpreting the presence of non-local resources at Vale Boi as the result of exchange and social networks, then we assume the existence of at least two maintained social networks, previously suggested by technological studies [focused mainly in the Proto-Solutrean and Solutrean technocomplexes, @belmiro_gravettian-solutrean_2021; @cascalheiraINFLUENCIAMEDITERRANICANAS].

However, the existence of long-distance mobility in hunter-gatherer groups during the UP is, in most cases, expected. Except for situations where specific technological choices may be interpreted as ethnographic and cultural boundaries between different groups, possibly indicating small territories, the very basis of a hunter-gatherer lifeway is marked by mobility and adaptation to the shifts in natural resources and the necessity to respect their renewal times [@kellyForagingSpectrumDiversity1995].

@@

Evidence from the site suggests that residential mobility patterns evolved throughout the UP. During the Gravettian, Vale Boi presents archaeological signals of short-term occupations, indicating higher mobility. This correlates with greater frequencies of non-local cherts, particularly those in the TL group, which may reflect either broader ranging mobility or individual provisioning strategies. In contrast, from the Proto-Solutrean onward, Vale Boi transitions into a longer-term residential settlement. Correspondingly, the frequency of non-local lithics declines, with local cherts becoming dominant during the Solutrean, consistent with reduced residential mobility [@surovellBehavioralEcologyLithic2009; @groveMovingFarMoving2023a].

*How were different chert types managed by hunter-gatherer groups, and what impact did mobility have on their technological organisation?*

These shifts in mobility frequency during the UP not only influenced the quantity of non-local raw materials found at Vale Boi but also shaped the technological and typological characteristics of the lithic assemblages. When focusing specifically on the relationship between occupation duration and technological behaviour, short-term occupations—associated with higher mobility—tend to correspond with increased frequencies of retouched tools, a pattern well-documented in ethnographic and archaeological literature [@torrenceTimeBudgetingHuntergatherer1983; @surovellBehavioralEcologyLithic2009; @groveMovingFarMoving2023a]. The more frequent occurrence of burins, for instance, may reflect an adaptive strategy to mobility, as these tools could serve as cores for the production of bladelets [@bartonGraverReconsideringBurin1996].

Across all UP phases at Vale Boi, regardless of occupation length, non-local cherts—costlier in terms of acquisition time and transport effort, even when acquired through exchange—consistently show higher proportions of retouched tools and distinct tool types. This pattern supports interpretations that non-local materials were not only more carefully curated but also reflect a higher investment in production and maintenance, aligning with theoretical expectations surrounding raw material economy [@kellyThreeSidesBiface1988; @perlesSearchLithicStrategies1992; @andrefskyRawMaterialAvailabilityOrganization1994; @kuhnUpperPaleolithicRaw2004].

The types of produced tool show diverse set of tool types, although less specialised tools, such as retouched blanks, remain predominant in non-local cherts across all assemblages. This suggests that these materials were not only transported as unspecialised tools and easily reconfigurable tools (e.g., retouched blanks) as suggested in previous studies [@odellMorphologicalExpressFunction1981; @siegelFunctionalVariabilityAssemblage1984; @shottTechnologicalOrganizationSettlement1986], but as volumes that supported tool specialisation and diversification.

In addition, non-local cherts consistently exhibit higher knapping intensity, indicating an increased effort to maximise their utility once transported to the site. This behaviour conforms to both theoretical models and archaeological expectations regarding the conservation and extended use of high-cost raw materials [@rothProductionTransportBlanks1998; @kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009].

This indicates that despite variations in residential mobility throughout the UP, the lithic record at Vale Boi reveals differentiated strategies of raw material management, where distance from source emerges as a key determinant in shaping technological investment and tool use.

*How did technological and social organisation evolve over approximately 10,000 years of the UP in southwestern Portugal, and how might climatic changes have influenced these adaptations?*

@@

After the Heinrich stadial, which once again created mild, mediterranean conditions all throughout Iberia, the onset of the Proto-Solutrean and, subsequently, the Solutrean, coincided with Heinrich Event 2 (HE 2) and the Last Glacial Maximum (LGM) [@belmiro_gravettian-solutrean_2021; @cascalheiraHunterGathererEcodynamics2013; @gonzalez-samperiz_steppes_2010-1; @schmidtRapidClimateChange2012]. It is also during this time frame [c. 26 and 20 ka cal BP, @belmiro_gravettian-solutrean_2021; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013] that the settlement patterns and raw material proportions begin to show alterations, after what may have been a small occupational hiatus before the Proto-Solutrean represented by lower frequencies of archaeological materials.

As such, we suggest that as climatic conditions worsened, transforming the southern Iberian landscape from Mediterranean forests to desert-like steppes [@gomezRefugiaRefugiaPatterns2007; @gonzalez-samperiz_steppes_2010-1; @hewitt_genetic_2000], hunter-gatherer groups continued seasonal mobility, but at a reduced frequency. According to Surovell’s [-@surovellBehavioralEcologyLithic2009] “Mean per Capita Occupation Span” model, this reduction likely stemmed from resource scarcity and the need to avoid overexploitation of ecological niches.

A similar pattern is observed in the UP occupation of the Côa Valley, where from the Proto-Solutrean onwards, the proportions of non-local raw materials decreased with an increased reliance on regional resources in comparison to the Gravettian occupations [@aubryWeWillBe2012]. This may indicate a similar behaviour across Iberia motivated by the occupation of ecological niches.

While it would then be expected that social networks would see themselves expanded during the HE 2 and LGM, given the scarcity of resources, and thus a higher presence of long-distance raw materials on Proto-Solutrean and Solutrean occupations, the existence of these ecological niches instead possibly fostered an environment of relative stability.
